---
title: "R WorkShop 2"
output:
  html_document:
    df_print: paged
---
#Tidyverse Functions: an overview

In this Document we will explore a number of functions in the tidyverse library collection
to assist you in manipulating data sets in efficient and easy to read ways.

The functions we will be looking at are:
- select()
- filter()
- group_by()
- tally()
- arrange()
- mutate()
- mutate_at()

The Data set we will use in our examples is called mtcars, and is a built in 
dataset with the R programming language. The dataset contains data on motor trends 
and road tests.

Format of the dataframe: 
A data frame with 32 observations(rows) on 11 (numeric) variables(columns).

[, 1]	mpg	Miles/(US) gallon
[, 2]	cyl	Number of cylinders
[, 3]	disp	Displacement (cu.in.)
[, 4]	hp	Gross horsepower
[, 5]	drat	Rear axle ratio
[, 6]	wt	Weight (1000 lbs)
[, 7]	qsec	1/4 mile time
[, 8]	vs	Engine (0 = V-shaped, 1 = straight)
[, 9]	am	Transmission (0 = automatic, 1 = manual)
[,10]	gear	Number of forward gears
[,11]	carb	Number of carburetors

###Load Libraries and Dataset
```{r, load Librarys, echo=FALSE, warning=FALSE, message=FALSE}
# Load the required library
library(tidyverse)
library(reactable)
# library(lubridate)  for %m+% months(3) example

# store our dataset mtcars in the variable 'data'
data <- mtcars

# Display the size of the dataset
dim(data)

# Display the dataset
data
```

###Select Specific Columns
The function Select() comes from the library called 'dplyr' which is a part of tidyverse.
This function allows us to subset a dataframe by specifying column names.
```{r}
# We are selecting columns mpg, disp, hp from the dataframe 'data' 
# and storing it in the new dataframe called 'subset1'
subset1 <- data %>% select(mpg, disp, hp)

# We are selecting columns drat, cyl, wt from the dataframe 'data' 
# and storing it in the new dataframe called 'subset2'
subset2 <- data %>% select(drat, cyl, wt)

# We are selecting columns mpg, hp from the dataframe 'subset1' 
# and storing it in the new dataframe called 'subset1.2'
subset1.2 <- subset1 %>% select(mpg, hp)

# We are displaying the three dataframes we've subsetted.
subset1
subset2
subset1.2
```


###Filter Rows by Cell Contents
```{r}
efficient_cars <- data %>% filter(mpg > 18)
efficient_cars

fast_cars <- data %>% filter(hp >200)
fast_cars

not_8_cyl <- data %>% filter(cyl!=8)
not_8_cyl 

automatic <- data %>% filter(am==0)
automatic

manual <- data %>% filter(am!=0)
# same as 
manual <- data %>% filter(am==1)
manual

# Combining Queries into a single statement

fast_manual <- data %>% filter((hp>200)&(am==1))
fast_manual

heavy_inefficient <- data %>% filter((wt>3.7)&(mpg<18))
heavy_inefficient
```


###Group_by
```{r}
# Grouping things wont change how they look, but will change how they behave
cylinders <- data %>% group_by(cyl)
cylinders

# returns the max weight for each cylinder group
max_wt <- cylinders %>% filter(wt == max(wt))
# returns the max weight for each cylinder group
min_wt <- cylinders %>% filter(wt == min(wt))

# merge the min and max weights by cylinder into a new df
min_max_wt_by_cyl <- rbind(max_wt, min_wt)
min_max_wt_by_cyl

# To remove the grouping attribute from a dataframe
cylinders %>% ungroup()
```
###Arrange
```{r}
# Descending horse power
sort_hp <- data %>% arrange(desc(hp))
sort_hp

# Ascending weight
weight_wt <- data %>% arrange(wt)
weight_wt

# Sorting while preserving groups
# These amount to the same thing, as the cylinders dataframe is already grouped
cylinders_sorted <- cylinders %>% arrange(wt, .by_group=TRUE)
cylinders_sorted <- data %>% group_by(cyl) %>% arrange(wt, .by_group=TRUE)
cylinders_sorted 

# arranging across multiple columns at once
# hp mpg
hp_by_mpg <- data %>% arrange(hp, mpg)
hp_by_mpg
```


###Tally
```{r}

am_count <- data %>% group_by(am) %>% tally() %>% arrange(desc(n))
am_count


cyl_count <- data %>% group_by(cyl) %>% tally(sort=TRUE)
cyl_count
```






###Mutate : change multiple cells
```{r}
# the spread of mpg on the dataset
mpg_range <- data %>% mutate(mpg_diff = mpg-(min(mpg))) 
mpg_range %>% select(mpg, mpg_diff) %>% arrange(mpg_diff)


# to create a proportions column on the manual/automatic tally am
total <- sum(am_count$n)
am_count <- am_count %>% mutate(proportion = ((n/total)))
am_count <- am_count %>% mutate(percentage = (signif((proportion), 2)*100))
am_count

```



###Mutate_at : change multiple cells
```{r}
# function to turn miles per gallon into kilometers per litre
mpg_2_kpl <- function(mpg){
  return(mpg*0.425144)
}
# function to turn pounds into kilograms
lbs_2_kg <- function(lbs){
  return(lbs*0.453592)
}

canadian_cars <- data %>% mutate_at(vars(mpg), mpg_2_kpl)
canadian_cars <- canadian_cars %>% mutate_at(vars(wt), lbs_2_kg)
canadian_cars


```





